Orienting the causal relationship between imprecisely measured traits using genetic instruments

### 2017-03-15 docs/manuscript.Rmd
updated

where $\hat{y} = \hat{\beta}_x x$. MR analysis estimates the causal influence of $x$ on $y$ by using the instrument as a proxy for $x$, such that

@@

We demonstrate this new method by evaluating the causal relationships of `r nrow(qtl_orig)` known associations between DNA methylation and gene expression levels using summary level data. The causal direction is heavily influenced by how much measurement error is present in the different assaying platforms. For example, if DNA methylation measures typically have higher measuremet error than gene expression measures then our analysis suggests that DNA methylation levels would be more often the causal factor in the association. Indeed, previous studies which have evaluated measurement error in these platforms do support this position [@Bose2014; @Bryant2011], though making strong conclusions for this analysis is difficult because measurement error is likely to be study specific.

In our simulations we focused on the simple case of a single instrument in a single sample setting with a view to making a fair comparison between MR and the various mediation-based methods available. We note that it is straightforward to extend the MR Steiger approach to multiple instruments, requiring only that the total variance explained by all instruments be calculated under the assumption that they are independent. Multiple instruments can indeed help to circumvent some of the problems described here in other ways also, for example by evaluating the proportionality of the SNP-exposure and SNP-outcome effects [@DaveySmithHemani2014]. Additionally, if there is at least one instrument for each trait then bi-directional MR can offer solutions to inferring the causal direction [@Richmond2014; @DaveySmithHemani2014; @Mancuso2017]. We restricted the simulations to evaluating the causal inference between quantitative traits, but it is possible that the analysis could be extended to binary traits by using the genetic variance explained on the liability scale, taking into account the population prevalence [@Lee2013c]. However, our analysis goes beyond many previous explorations of measurement error by assessing the impacts of both imprecision (noise) and linear transformations of the true variable on causal inference.

@@

>>>>

where $\hat{y} = \hat{\beta}_x x$ and assuming no intercept $y - \hat{y} = \epsilon_x$. MR analysis estimates the causal influence of $x$ on $y$ by using the instrument as a proxy for $x$, such that

@@

We demonstrate this new method by evaluating the causal relationships of `r nrow(qtl_orig)` known associations between DNA methylation and gene expression levels using summary level data. The inferred causal direction is heavily influenced by how much measurement error is present in the different assaying platforms. For example, if DNA methylation measures typically have higher measuremet error than gene expression measures then our analysis suggests that DNA methylation levels would be more often the causal factor in the association. Indeed, previous studies which have evaluated measurement error in these platforms do support this position [@Bose2014; @Bryant2011], though making strong conclusions for this analysis is difficult because measurement error is likely to be study specific. We also haven't accounted for the influence of winner's curse, which can inflate estimates of the variance explained by SNPs, with higher inflation expected amongst lower powered studies. Using p-values for genetic associations from replication studies will mitigate this problem.

In our simulations we focused on the simple case of a single instrument in a single sample setting with a view to making a fair comparison between MR and the various mediation-based methods available. However, if there is only a single instrument it is difficult to separate between the two competing models of $g$ instrumenting a trait which causes another trait, and $g$ having pleiotropic effects on both traits independently [@Zhu2016]. Under certain conditions of measurement error the CIT test can distinguish these models. We also note that it is straightforward to extend the MR Steiger approach to multiple instruments, requiring only that the total variance explained by all instruments be calculated under the assumption that they are independent. Multiple instruments can indeed help to distinguish between the causal and pleiotropic models, for example by evaluating the proportionality of the SNP-exposure and SNP-outcome effects [@DaveySmithHemani2014]. Additionally, if there is at least one instrument for each trait then bi-directional MR can offer solutions to inferring the causal direction [@Richmond2014; @DaveySmithHemani2014; @Mancuso2017]. We restricted the simulations to evaluating the causal inference between quantitative traits, but it is possible that the analysis could be extended to binary traits by using the genetic variance explained on the liability scale, taking into account the population prevalence [@Lee2013c]. However, our analysis goes beyond many previous explorations of measurement error by assessing the impacts of both imprecision (noise) and linear transformations of the true variable on causal inference.

@@

## Acknowledgements

This work was funded by the UK Medical Research Council (MC_UU_12013/1 and MC_UU_12013/9).

### 2017-03-15 docs/manuscript.tex
updated

\setlength{\droptitle}{-2em}
  \title{Orienting the causal relationship between imprecisely measured traits
using genetic instruments}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{13 March 2017}

@@

We demonstrate this new method by evaluating the causal relationships of
458 known associations between DNA methylation and gene expression
levels using summary level data. The causal direction is heavily
influenced by how much measurement error is present in the different
assaying platforms. For example, if DNA methylation measures typically
have higher measuremet error than gene expression measures then our
analysis suggests that DNA methylation levels would be more often the
causal factor in the association. Indeed, previous studies which have
evaluated measurement error in these platforms do support this position
(45,46), though making strong conclusions for this analysis is difficult
because measurement error is likely to be study specific. We also
haven't accounted for the influence of winner's curse, which can inflate
estimates of the variance explained by SNPs, with higher inflation
expected amongst lower powered studies. Using p-values for genetic
associations from replication studies will mitigate this problem.

In our simulations we focused on the simple case of a single instrument
in a single sample setting with a view to making a fair comparison
between MR and the various mediation-based methods available. We note
that it is straightforward to extend the MR Steiger approach to multiple
instruments, requiring only that the total variance explained by all
instruments be calculated under the assumption that they are
independent. Multiple instruments can indeed help to circumvent some of
the problems described here in other ways also, for example by
evaluating the proportionality of the SNP-exposure and SNP-outcome
effects (16). Additionally, if there is at least one instrument for each
trait then bi-directional MR can offer solutions to inferring the causal
direction (16,47,48). We restricted the simulations to evaluating the
causal inference between quantitative traits, but it is possible that
the analysis could be extended to binary traits by using the genetic
variance explained on the liability scale, taking into account the
population prevalence (49). However, our analysis goes beyond many
previous explorations of measurement error by assessing the impacts of
both imprecision (noise) and linear transformations of the true variable
on causal inference.

In this work we assumed that pleiotropy (the influence of the instrument
on the outcome through a mechanism other than the exposure) was not
present. Recent method developments in MR (24,25) have focused on
accounting for the issues that horizontal pleiotropy can introduce when
multiple instruments are available, but how they perform in the presence
of measurement error remains to be explored. An important advantage that
MR confers over most mediation based analysis is that it can be
performed in two samples, which can considerably improve power and
expand the scope of analysis. However, whether there is a substantive
difference in two sample MR versus one sample MR in how measurement
error has an effect is not yet fully understood. We have also assumed no
measurement error in the genetic instrument, which is not unreasonable
given the strict QC protocols that ensure high quality genotype data is
available to most studies. We have restricted the scope to only
exploring non-differential measurement error and avoided the
complications incurred if measurement error in the exposure and outcome
is correlated. We have also not addressed other issues pertaining to
instrumental variables which are relevant to the question of
instrument-exposure specification. One such problem is exposure
misspecification, for example an instrument could associate with several
closely related putative outcomes, with only one of them actually having
a causal effect on the outcome. This problem has shown to be the case
for SNPs influencing different lipid fractions, for example (50,51).

Mediation based network approaches, that go beyond analyses of two
variables, are very well established (35) and have a number of
extensions that make them valuable tools, including for example network
construction. But because they are predicated on the basic underlying
principles of mediation they are liable to suffer from the same issues
of measurement error. Recent advances in MR methodology, for example
applying MR to genetical genomics (52), multivariate MR (51) and
mediation through MR (53--55) may offer more robust alternatives for
these more complicated problems.

@@

@@

\hypertarget{ref-Millstein2016}{}
18. Millstein J, Chen GK, Breton CV. Cit : Hypothesis Testing Software
for Mediation Analysis in Genomic Applications. Bioinformatics.
2016;AOP(10.1093/bioinformatics/btw135):1--10.

@@

\hypertarget{ref-Richmond2014}{}
47. Richmond RC, Davey Smith G, Ness AR, Hoed M den, McMahon G, Timpson
NJ. Assessing Causality in the Association between Child Adiposity and
Physical Activity Levels: A Mendelian Randomization Analysis. Ludwig DS,
editor. PLoS Medicine {[}Internet{]}. 2014 Mar;11(3):e1001618. Available
from: \url{http://dx.plos.org/10.1371/journal.pmed.1001618}

\hypertarget{ref-Mancuso2017}{}
48. Mancuso N, Shi H, Goddard P, Kichaev G, Gusev A, Pasaniuc B.
Integrating Gene Expression with Summary Association Statistics to
Identify Genes Associated with 30 Complex Traits. The American Journal
of Human Genetics {[}Internet{]}. 2017 Mar;100(3):473--87. Available
from:
\url{http://linkinghub.elsevier.com/retrieve/pii/S0002929717300320}

\hypertarget{ref-Lee2013c}{}
49. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
association studies: quantification of power and genomic prediction
accuracy. PLoS One. 2013;8(8):e71494.

\hypertarget{ref-Do2013}{}
50. Do R, Willer CJ, Schmidt EM, Sengupta S, Gao C, Peloso GM, et al.
Common variants associated with plasma triglycerides and risk for
coronary artery disease. Nature Genetics {[}Internet{]}. 2013
Oct;45(11):1345--52. Available from:
\url{http://www.nature.com/doifinder/10.1038/ng.2795}

\hypertarget{ref-Burgess2014a}{}
51. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
multivariable Mendelian randomization to disentangle the causal effects
of lipid fractions. PloS one {[}Internet{]}. 2014 Jan;9(10):e108891.
Available from:
\url{http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0108891}

\hypertarget{ref-Relton2012}{}
52. Relton CL, Davey Smith G. Two-step epigenetic Mendelian
randomization: a strategy for establishing the causal role of epigenetic
processes in pathways to disease. International journal of epidemiology
{[}Internet{]}. 2012 Feb;41(1):161--76. Available from:
\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3304531\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3304531\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}

\hypertarget{ref-Varbo2015}{}
53. Varbo A, Benn M, Smith GD, Timpson NJ, Tybjaerg-Hansen A,
Nordestgaard BG. Remnant cholesterol, low-density lipoprotein
cholesterol, and blood pressure as mediators from obesity to ischemic
heart disease. Circulation research {[}Internet{]}. 2015
Feb;116(4):665--73. Available from:
\url{http://www.ncbi.nlm.nih.gov/pubmed/25411050}

\hypertarget{ref-Burgess2015}{}
54. Burgess S, Daniel RM, Butterworth AS, Thompson SG. Network Mendelian
randomization: using genetic variants as instrumental variables to
investigate mediation in causal pathways. International journal of
epidemiology {[}Internet{]}. 2015 Apr;44(2):484--95. Available from:
\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=4469795\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=4469795\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}

\hypertarget{ref-Richmond2016}{}
55. Richmond RC, Hemani G, Tilling K, Davey Smith G, Relton CL.
Challenges and novel approaches for investigating molecular mediation.
Human molecular genetics {[}Internet{]}. 2016 Oct;25(R2):R149--56.
Available from:
\href{http://www.ncbi.nlm.nih.gov/pubmed/27439390\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC5036871}{http://www.ncbi.nlm.nih.gov/pubmed/27439390 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC5036871}

>>>>

\setlength{\droptitle}{-2em}
  \title{Orienting the causal relationship between imprecisely measured traits
using genetic instruments}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{15 March 2017}

@@

We demonstrate this new method by evaluating the causal relationships of
458 known associations between DNA methylation and gene expression
levels using summary level data. The inferred causal direction is
heavily influenced by how much measurement error is present in the
different assaying platforms. For example, if DNA methylation measures
typically have higher measuremet error than gene expression measures
then our analysis suggests that DNA methylation levels would be more
often the causal factor in the association. Indeed, previous studies
which have evaluated measurement error in these platforms do support
this position (45,46), though making strong conclusions for this
analysis is difficult because measurement error is likely to be study
specific. We also haven't accounted for the influence of winner's curse,
which can inflate estimates of the variance explained by SNPs, with
higher inflation expected amongst lower powered studies. Using p-values
for genetic associations from replication studies will mitigate this
problem.

In our simulations we focused on the simple case of a single instrument
in a single sample setting with a view to making a fair comparison
between MR and the various mediation-based methods available. However,
if there is only a single instrument it is difficult to separate between
the two competing models of \(g\) instrumenting a trait which causes
another trait, and \(g\) having pleiotropic effects on both traits
independently (47). Under certain conditions of measurement error the
CIT test can distinguish these models. We also note that it is
straightforward to extend the MR Steiger approach to multiple
instruments, requiring only that the total variance explained by all
instruments be calculated under the assumption that they are
independent. Multiple instruments can indeed help to distinguish between
the causal and pleiotropic models, for example by evaluating the
proportionality of the SNP-exposure and SNP-outcome effects (16).
Additionally, if there is at least one instrument for each trait then
bi-directional MR can offer solutions to inferring the causal direction
(16,48,49). We restricted the simulations to evaluating the causal
inference between quantitative traits, but it is possible that the
analysis could be extended to binary traits by using the genetic
variance explained on the liability scale, taking into account the
population prevalence (50). However, our analysis goes beyond many
previous explorations of measurement error by assessing the impacts of
both imprecision (noise) and linear transformations of the true variable
on causal inference.

In this work we assumed that pleiotropy (the influence of the instrument
on the outcome through a mechanism other than the exposure) was not
present. Recent method developments in MR (24,25) have focused on
accounting for the issues that horizontal pleiotropy can introduce when
multiple instruments are available, but how they perform in the presence
of measurement error remains to be explored. An important advantage that
MR confers over most mediation based analysis is that it can be
performed in two samples, which can considerably improve power and
expand the scope of analysis. However, whether there is a substantive
difference in two sample MR versus one sample MR in how measurement
error has an effect is not yet fully understood. We have also assumed no
measurement error in the genetic instrument, which is not unreasonable
given the strict QC protocols that ensure high quality genotype data is
available to most studies. We have restricted the scope to only
exploring non-differential measurement error and avoided the
complications incurred if measurement error in the exposure and outcome
is correlated. We have also not addressed other issues pertaining to
instrumental variables which are relevant to the question of
instrument-exposure specification. One such problem is exposure
misspecification, for example an instrument could associate with several
closely related putative outcomes, with only one of them actually having
a causal effect on the outcome. This problem has shown to be the case
for SNPs influencing different lipid fractions, for example (51,52).

Mediation based network approaches, that go beyond analyses of two
variables, are very well established (35) and have a number of
extensions that make them valuable tools, including for example network
construction. But because they are predicated on the basic underlying
principles of mediation they are liable to suffer from the same issues
of measurement error. Recent advances in MR methodology, for example
applying MR to genetical genomics (53), multivariate MR (52) and
mediation through MR (54--56) may offer more robust alternatives for
these more complicated problems.

@@

\subsection{Acknowledgements}\label{acknowledgements}

This work was funded by the UK Medical Research Council (MC\_UU\_12013/1
and MC\_UU\_12013/9).

@@

\hypertarget{ref-Millstein2016}{}
18. Millstein J. cit: Causal Inference Test. R package version 1.9
{[}Internet{]}. 2016. Available from:
\url{http://cran.r-project.org/package=cit}

@@

\hypertarget{ref-Zhu2016}{}
47. Zhu Z, Zhang F, Hu H, Bakshi A, Robinson MR, Powell JE, et al.
Integration of summary data from GWAS and eQTL studies predicts complex
trait gene targets. Nature Genetics {[}Internet{]}. 2016
Mar;48(5):481--7. Available from:
\url{http://www.nature.com/doifinder/10.1038/ng.3538}

\hypertarget{ref-Richmond2014}{}
48. Richmond RC, Davey Smith G, Ness AR, Hoed M den, McMahon G, Timpson
NJ. Assessing Causality in the Association between Child Adiposity and
Physical Activity Levels: A Mendelian Randomization Analysis. Ludwig DS,
editor. PLoS Medicine {[}Internet{]}. 2014 Mar;11(3):e1001618. Available
from: \url{http://dx.plos.org/10.1371/journal.pmed.1001618}

\hypertarget{ref-Mancuso2017}{}
49. Mancuso N, Shi H, Goddard P, Kichaev G, Gusev A, Pasaniuc B.
Integrating Gene Expression with Summary Association Statistics to
Identify Genes Associated with 30 Complex Traits. The American Journal
of Human Genetics {[}Internet{]}. 2017 Mar;100(3):473--87. Available
from:
\url{http://linkinghub.elsevier.com/retrieve/pii/S0002929717300320}

\hypertarget{ref-Lee2013c}{}
50. Lee SH, Wray NR. Novel genetic analysis for case-control genome-wide
association studies: quantification of power and genomic prediction
accuracy. PLoS One. 2013;8(8):e71494.

\hypertarget{ref-Do2013}{}
51. Do R, Willer CJ, Schmidt EM, Sengupta S, Gao C, Peloso GM, et al.
Common variants associated with plasma triglycerides and risk for
coronary artery disease. Nature Genetics {[}Internet{]}. 2013
Oct;45(11):1345--52. Available from:
\url{http://www.nature.com/doifinder/10.1038/ng.2795}

\hypertarget{ref-Burgess2014a}{}
52. Burgess S, Freitag DF, Khan H, Gorman DN, Thompson SG. Using
multivariable Mendelian randomization to disentangle the causal effects
of lipid fractions. PloS one {[}Internet{]}. 2014 Jan;9(10):e108891.
Available from:
\url{http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0108891}

\hypertarget{ref-Relton2012}{}
53. Relton CL, Davey Smith G. Two-step epigenetic Mendelian
randomization: a strategy for establishing the causal role of epigenetic
processes in pathways to disease. International journal of epidemiology
{[}Internet{]}. 2012 Feb;41(1):161--76. Available from:
\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3304531\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3304531\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}

\hypertarget{ref-Varbo2015}{}
54. Varbo A, Benn M, Smith GD, Timpson NJ, Tybjaerg-Hansen A,
Nordestgaard BG. Remnant cholesterol, low-density lipoprotein
cholesterol, and blood pressure as mediators from obesity to ischemic
heart disease. Circulation research {[}Internet{]}. 2015
Feb;116(4):665--73. Available from:
\url{http://www.ncbi.nlm.nih.gov/pubmed/25411050}

\hypertarget{ref-Burgess2015}{}
55. Burgess S, Daniel RM, Butterworth AS, Thompson SG. Network Mendelian
randomization: using genetic variants as instrumental variables to
investigate mediation in causal pathways. International journal of
epidemiology {[}Internet{]}. 2015 Apr;44(2):484--95. Available from:
\href{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=4469795\%7B/\&\%7Dtool=pmcentrez\%7B/\&\%7Drendertype=abstract}{http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=4469795\{\textbackslash{}\&\}tool=pmcentrez\{\textbackslash{}\&\}rendertype=abstract}

\hypertarget{ref-Richmond2016}{}
56. Richmond RC, Hemani G, Tilling K, Davey Smith G, Relton CL.
Challenges and novel approaches for investigating molecular mediation.
Human molecular genetics {[}Internet{]}. 2016 Oct;25(R2):R149--56.
Available from:
\href{http://www.ncbi.nlm.nih.gov/pubmed/27439390\%20http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC5036871}{http://www.ncbi.nlm.nih.gov/pubmed/27439390 http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=PMC5036871}
